---
title: "Categorical variable functions"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

The **pols503** package contains a couple functions designed to facilitate data cleaning with categorical variables while using `%>%`.

This vignette uses the following packages:
```{r message=FALSE}
library("pols503")
library("gapminder")
library("tidyr")
library("dplyr")
```

## Dummies

The function `to_dummies` converts a character or factor variable to dummy variables. 
The function `from_dummies` converts a set of dummy variables to a character vector.

For this example, we will use the `gapminder` data:
```{r}
data("gapminder")
```
The variable `continent` indicates the continent of each variable. 
```{r}
glimpse(gapminder)
```
To convert `continent` to a set of dummy variables.
```{r}
gapminder2 <- to_dummies(gapminder, continent, sep="_")
glimpse(gapminder2)
```
Intead of a single variable `contintent`, the dataset `gapminder2` contains a set of dummy variables named `gapminder_*`, one for each value of the original variable, `continent`.
To avoid dropping the original variable, use the argument `remove = FALSE`.

Now that you have a set of dummy variables, perhaps you would like to convert them back to a categorical variable.
```{r}
gapminder3 <- from_dummies(gapminder2, continent, matches("^continent_"))
glimpse(gapminder3)
```
Now `gapminder3` contains a character vector containing the names of the dummy vectors.
The categories of `continent` are all prefixed with "continent_", but this can be cleaned up using functions from **tidyr**:
```{r}
gapminder3 %>%
  separate(continent, c("continent_prefix", "continent")) %>%
  select(- continent_prefix)
```

## Categorize

The function `categorize` creates a character vector from a set of logical expressions.
The following code creates categories for low life expectancy for years 1980 and before, and years after 1980, with the default category being "high lifeExp"
```{r}
categorize(gapminder, lifeExpCat,
           "low lifeExp, pre 1980" = year <= 1980 & lifeExp < 55,
            "low lifeExp, pre 1980" = year > 1980 & lifeExp < 60,
           .default = "high lifeExp") %>%
  glimpse()
```



